---
title: "RNAseq Analysis"
output: html_document
date: "2025-04-09"
---

# Setup and load libraries
```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(clusterProfiler)
library(org.Hs.eg.db)
library(enrichplot)
library(tximport)
library(DESeq2)
library(ggplot2)
library(pheatmap)
library(ComplexHeatmap)
library(EnhancedVolcano)
library(RColorBrewer)
library(dplyr)
library(tibble)
library(ggrepel)
library(gplots)
library(reshape2)
library(RNAseqQC)
library(purrr)
library(tidyr)
library(magrittr)
library(circlize)
```

# Load and format data
```{r load-data}
# txi$counts should contain gene-level data
counts <- read.csv("annotated_tmm_normalized_counts.csv")

counts <- counts[!duplicated(counts$name), ] # Remove duplicate gene name rows
counts <- na.omit(counts) # Remove NA/missing values
rownames(counts) <- counts$name # Set rownames
counts <- counts[ , -(1:3)] # Remove first 3 columns
counts <- round(counts) # DESeq2 expects integer counts

# Filter counts df
counts <- counts[rowSums(counts) >= 10, ]

# Sample metadata
sample_table <- data.frame(
  sampleName = colnames(counts),
  knockout_status = c("WT", "WT", "ZFP36L1 -/-", "ZFP36L1 -/-", "ZFP36L1 -/-"),
  bisphenol = c("BPA", "BPF", "Untreated", "BPA", "BPF")
)

rownames(sample_table) <- sample_table$sampleName
```

# Boxplot of raw counts
```{r boxplot-tmm}
# Log transform for visualization
log_counts <- log2(counts + 1)

# Boxplot
boxplot(log_counts, las = 2, col = "white", main = "Boxplot of TMM Counts (log2)")
```

# Create DESeq object
```{r dds-setup}
dds <- DESeqDataSetFromMatrix(
  countData = counts,
  colData = sample_table,
  design = ~ knockout_status + bisphenol
)

# Prefilter low counts
keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep, ]

dds
```

# QC plots on DESeq object
```{r qc-plots}
# Library complexity
plot_library_complexity(dds)

# Gene detection
plot_gene_detection(dds)
```

# Variance stabilizations
```{r var-stab}
vsd <- vst(dds)
mean_sd_plot(vsd)
```

# Replicate/sample variability
```{r rep-var}
# define new grouping variable
colData(vsd)$trt_data <- paste0(colData(vsd)$bisphenol, "_", colData(vsd)$kockout_status)

ma_plots <- plot_sample_MAs(vsd, group = "trt_data")
cowplot::plot_grid(plotlist = ma_plots[17:24], ncol = 2)
```

# Sample clustering
```{r sample-clust}
# set seed to control random annotation colors
set.seed(123)
plot_sample_clustering(vsd, anno_vars = c("bisphenol", "knockout_status"), distance = "euclidean")
```

# Principal component analysis (PCA)
```{r pca}
plot_pca(vsd, PC_x = 1, PC_y = 2, color_by = "knockout_status", shape_by = "bisphenol")
```

# PCA loadings
```{r pca-loadings}
pca_res <- plot_pca(vsd, show_plot = FALSE)
plot_loadings(pca_res, PC = 1, annotate_top_n = 5)
```

# Investigate multiple PCAs
```{r investigate-pca}
plot_pca_scatters(vsd, n_PCs = 4, color_by = "knockout_status", shape_by = "bisphenol", show_var_exp = FALSE)
```

# Differential testing
```{r diff-test}
# design variables need to be factors
dds$knockout_status <- as.factor(dds$knockout_status)
dds$bisphenol <- as.factor(dds$bisphenol)
design(dds) <- ~ knockout_status + bisphenol

dds <- DESeq(dds, parallel = T)
plotDispEsts(dds, legend = FALSE) # black = gene-est, red = fitted, blue = final
```

# Write CSV file of differential testing results
```{r save-de-res}
# KO vs WT
# log2FoldChange > 0 means gene is higher in KO than WT
# log2FoldChange < 0 means gene is lower in KO than WT
res_ko_vs_wt <- results(dds, contrast = c("knockout_status", "ZFP36L1 -/-", "WT"))
res_ko_vs_wt$gene <- rownames(res_ko_vs_wt)
write.csv(as.data.frame(res_ko_vs_wt), file = "DEA_ZFP36L1ko_vs_WT.csv", row.names = FALSE)

# BPA vs Untreated
# log2FoldChange > 0 means gene is higher in BPA than UN
# log2FoldChange < 0 means gene is lower in BPA than UN
res_bpa_vs_ctrl <- results(dds, contrast = c("bisphenol", "BPA", "Untreated"))
res_bpa_vs_ctrl$gene <- rownames(res_bpa_vs_ctrl)
write.csv(as.data.frame(res_bpa_vs_ctrl), file = "DEA_BPA_vs_Untreated.csv", row.names = FALSE)

# BPF vs Untreated
# log2FoldChange > 0 means gene is higher in BPF than UN
# log2FoldChange < 0 means gene is lower in BPF than UN
res_bpf_vs_ctrl <- results(dds, contrast = c("bisphenol", "BPF", "Untreated"))
res_bpf_vs_ctrl$gene <- rownames(res_bpf_vs_ctrl)
write.csv(as.data.frame(res_bpf_vs_ctrl), file = "DEA_BPF_vs_Untreated.csv", row.names = FALSE)

# BPA vs BPF
# log2FoldChange > 0 means gene is higher in BPA than BPF
# log2FoldChange < 0 means gene is lower in BPA than BPF
res_bpa_vs_bpf <- results(dds, contrast = c("bisphenol", "BPA", "BPF"))
res_bpa_vs_bpf$gene <- rownames(res_bpa_vs_bpf)
write.csv(as.data.frame(res_bpa_vs_bpf), file = "DEA_BPA_vs_BPF.csv", row.names = FALSE)

# Could not run comparison between bisphenol treatment across conditions. This comparison caused issues as there is a high amount of collinearity and linear relationship between treatment when considering knockout status. This basically means that knockout status is what is driving any sort of difference and the bisphenols are having little to no impact on the gene expression. This can be seen in the PCA as well.
```

# Volcano plots
```{r volcano-plots}
EnhancedVolcano(res_ko_vs_wt,
    lab = res_ko_vs_wt$gene,
    x = 'log2FoldChange',
    y = 'padj',
    title = 'KO vs WT',
    pCutoff = 0.05,
    FCcutoff = 1,
    pointSize = 2.5,
    labSize = 3.5,
    max.overlaps = 10,
    legendPosition = 'right'
)

EnhancedVolcano(res_bpa_vs_ctrl,
    lab = res_bpa_vs_ctrl$gene,
    x = 'log2FoldChange',
    y = 'padj',
    title = 'BPA vs Untreated',
    pCutoff = 0.05,
    FCcutoff = 1,
    pointSize = 2.5,
    labSize = 3.5,
    max.overlaps = 10
)

EnhancedVolcano(res_bpf_vs_ctrl,
    lab = res_bpf_vs_ctrl$gene,
    x = 'log2FoldChange',
    y = 'padj',
    title = 'BPF vs Untreated',
    pCutoff = 0.05,
    FCcutoff = 1,
    pointSize = 2.5,
    labSize = 3.5
)

EnhancedVolcano(res_bpa_vs_bpf,
    lab = res_bpa_vs_bpf$gene,
    x = 'log2FoldChange',
    y = 'padj',
    title = 'BPA vs BPF',
    pCutoff = 0.05,
    FCcutoff = 1,
    pointSize = 2.5,
    labSize = 3.5
)
```

# Heatmap
```{r heatmap}
vsd <- vst(dds, blind = FALSE)  # or use rlog(dds, blind = FALSE)
vst_mat <- assay(vsd)  # This contains all genes × samples

vst_scaled <- t(scale(t(vst_mat)))  # Z-score normalization per gene

# Create annotation from metadata
metadata <- as.data.frame(colData(vsd)[, c("knockout_status", "bisphenol")])

# Define annotation colors
ann_colors <- list(
  knockout_status = c("WT" = "#1f77b4", "ZFP36L1 -/-" = "#d62728"),
  bisphenol = c("BPA" = "#ff7f0e", "BPF" = "#2ca02c", "Untreated" = "#9467bd")
)

# Create ComplexHeatmap annotation
column_ha <- HeatmapAnnotation(
  df = metadata,
  col = ann_colors,
  show_annotation_name = TRUE
)

Heatmap(
  vst_scaled,  # or vst_mat if not scaled
  name = "Z-score",
  top_annotation = column_ha,
  show_row_names = FALSE,       # Hide row names for readability
  show_column_names = TRUE,
  cluster_columns = TRUE,
  cluster_rows = TRUE,
  col = colorRamp2(c(-2, 0, 2), c("blue", "white", "red")),
  column_title = "Heatmap of All Genes",
  row_names_gp = gpar(fontsize = 6),
  column_names_gp = gpar(fontsize = 10)
)
```

# Functional enrichment analysis
## Prepare gene list
```{r prep-gene-list}
# Ensure the results are ordered by adjusted p-value
res_ko_vs_wt <- res_ko_vs_wt[order(res_ko_vs_wt$padj), ]

# Extract the gene symbols and log2 fold changes
gene_list <- res_ko_vs_wt$log2FoldChange
names(gene_list) <- res_ko_vs_wt$gene

# Remove NA values
gene_list <- na.omit(gene_list)

# Sort the gene list in decreasing order (required for GSEA)
gene_list <- sort(gene_list, decreasing = TRUE)
```

## Perform gene ontology (GO) enrichment analysis
```{r go-enr}
# Define the universe of genes (all genes tested)
universe_genes <- rownames(res_ko_vs_wt)

# Perform ORA for Biological Process (BP)
ego_bp <- enrichGO(
  gene          = names(gene_list),
  universe      = universe_genes,
  OrgDb         = org.Hs.eg.db,
  keyType       = "SYMBOL",
  ont           = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.05
)

# Perform ORA for Molecular Function (MF)
ego_mf <- enrichGO(
  gene          = names(gene_list),
  universe      = universe_genes,
  OrgDb         = org.Hs.eg.db,
  keyType       = "SYMBOL",
  ont           = "MF",
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.05
)

# Perform ORA for Cellular Component (CC)
ego_cc <- enrichGO(
  gene          = names(gene_list),
  universe      = universe_genes,
  OrgDb         = org.Hs.eg.db,
  keyType       = "SYMBOL",
  ont           = "CC",
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.05
)
```

## Gene set enrichment analysis (GSEA)
```{r gsea}
# GSEA for Biological Process (BP)
gsea_bp <- gseGO(
  geneList      = gene_list,
  OrgDb         = org.Hs.eg.db,
  keyType       = "SYMBOL",
  ont           = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.05,
  verbose       = FALSE
)

# GSEA for Molecular Function (MF)
gsea_mf <- gseGO(
  geneList      = gene_list,
  OrgDb         = org.Hs.eg.db,
  keyType       = "SYMBOL",
  ont           = "MF",
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.05,
  verbose       = FALSE
)

# GSEA for Cellular Component (CC)
gsea_cc <- gseGO(
  geneList      = gene_list,
  OrgDb         = org.Hs.eg.db,
  keyType       = "SYMBOL",
  ont           = "CC",
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.05,
  verbose       = FALSE
)
```

## KEGG Over-Representation Analysis (ORA)
```{r kegg-ora}
# Convert gene symbols to Entrez IDs
gene_symbols <- names(gene_list)
entrez_ids <- mapIds(org.Hs.eg.db, keys = gene_symbols, column = "ENTREZID", keytype = "SYMBOL", multiVals = "first")
entrez_ids <- na.omit(entrez_ids)

# Perform KEGG ORA
kegg_ora <- enrichKEGG(
  gene          = entrez_ids,
  organism      = 'hsa',
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.05
)
```

## KEGG GSEA
```{r kegg-gsea}
# Prepare the ranked list with Entrez IDs
gene_list_entrez <- gene_list[names(entrez_ids)]
names(gene_list_entrez) <- entrez_ids

# Perform KEGG GSEA
kegg_gsea <- gseKEGG(
  geneList      = gene_list_entrez,
  organism      = 'hsa',
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.05,
  verbose       = FALSE
)
```

## Visualize functional enrichment analysis results
```{r viz-enr-res}
# Dotplots of GO GSEA results
dotplot(gsea_mf, showCategory = 10, title = "GO MF GSEA", font.size = 12)
dotplot(gsea_cc, showCategory = 10, title = "GO CC GSEA", font.size = 12)

# KEGG ORA dotplot
dotplot(kegg_ora, showCategory = 10, title = "KEGG Pathway ORA", font.size = 12)

# KEGG GSEA enrichment map and dotplot
dotplot(kegg_gsea, showCategory = 10, title = "KEGG Pathway GSEA", font.size = 12)

# GSEA plot for a specific KEGG term
gseaplot2(kegg_gsea, geneSetID = kegg_gsea$ID[1], title = kegg_gsea$Description[1])
gseaplot2(kegg_gsea, geneSetID = kegg_gsea$ID[2], title = kegg_gsea$Description[2])
```

```{r}
sessionInfo()
```
